Generalized Bradley-Terry models

### 2025-04-01 inst/jss_hyper3_real.Rnw
capitalizations

    H[c(Celtic = lambda             )] %<>%      add(     76)
    H[c(                 Rangers = 1)] %<>%      add(46     )
    H[c(Celtic = lambda, Rangers = 1)] %<>% subtract(46 + 76)
    return(H)
}
@
%
Above we see package idiom for creating a log-likelihood function.  We
first create an empty {\tt hyper3} object with an appropriate names
attribute.  The next three lines, using \pkg{magrittr} assignment
pipes, give the results for Rangers at home, and the next three lines
give the result for Celtic at home.  It is then easy to create a
bespoke log-likelihood function for Rangers-Celtic matches:
%
<<useoldfirm, cache = TRUE>>=
oldfirm(1.88)
@
%
We see a log-likelihood function for all $82+42+76+46=245$
Rangers-Celtic matches that were not score draws, conditional on
assumed value $\lambda=1.88$.  This corresponds to the relevant terms
of Equation~\ref{firstfewRCFL}.  We may make inferences about the true
value of $\lambda$ using profile support:
%
<<profilelikeoldfirm, cache = TRUE>>=
f <- function(lambda){maxp(oldfirm(lambda), give = 1)$`Log-likelihood`}
lam <- seq(from = 0.9, to = 3, len = howmany)
logLike <- sapply(lam, f)
@
%
Thus \fct{f} returns the support at the evaluate, conditional on a
supplied value of $\lambda$.  We may use this to create a profile
likelihood (figure~\ref{profilesupportlambda}), or to find the maximum
likelihood estimate for $\lambda$:

@@

<<testlambdazero, cache = TRUE, eval = TRUE, echo = FALSE>>=
jj <- optimize(f = function(D){f(c(lambda=1, D))}, interval = c(0.1, 0.4), maximum = TRUE)
@
%
Further, we may calculate a support function over the
$(\lambda,D)$-plane, shown in figure~\ref{contourlambdaD}.  For
example, we may test the hypothesis that $\lambda=1$ [corresponding to
  the left hand side of figure~\ref{contourlambdaD}] for which the
support is about \Sexpr{round(f(maxv)-jj$objective,2)}, far
exceeding the two units of support per degree of freedom
criteron~\citep{edwards1992}.  Alternatively, we can observe that this
lies in the tail region of its asymptotic null $\chi^2_2$ distribution
with a $p$~value of about \Sexpr{signif(pchisq(f(maxv)-jj$objective,lower.tail=FALSE,df=2),digits=2)}.

@@

As an example, consider figure~\ref{tikzabcde}.  This shows a partial
probability tree diagram for some of the $5!=120$ possible order
statistics for five competitors.  The standard Plackett-Luce
likelihoods have been modified to account for two groups of mutually
supporting runners: $\left\lbrace a,b,c\right\rbrace$ with support
term $\lambda$, and $\left\lbrace d,e\right\rbrace$ with support term
$\mu$.  From START, the first runner to cross the finishing line has
standard Plackett-Luce probability.  Taking the top path as an
example, we see that the likelihood function for $a\succ b\succ c\succ
d\succ e$ would be

>>>>

    H[c(Celtic = lambda             )] %<>%      add(     76)
    H[c(                 Rangers = 1)] %<>%      add(46     )
    H[c(Celtic = lambda, Rangers = 1)] %<>% subtract(46 + 76)
    return(H)
}
@
%
Above we see package idiom for creating a log-likelihood function.  We
first create an empty {\tt hyper3} object with an appropriate names
attribute.  The next three lines, using \pkg{magrittr} assignment
pipes, give the results for Rangers at home, and the next three lines
give the result for Celtic at home.  It is then easy to create a
bespoke log-likelihood function for Rangers-Celtic matches:
%
<<useoldfirm, cache = TRUE>>=
oldfirm(1.88)
@
%
We see a log-likelihood function for all $82+42+76+46=245$
Rangers-Celtic matches that were not score draws, conditional on
assumed value $\lambda=1.88$.  This corresponds to the relevant terms
of Equation~\ref{firstfewRCFL}.  We may make inferences about the true
value of $\lambda$ using profile support:
%
<<profilelikeoldfirm, cache = TRUE>>=
f <- function(lambda){maxp(oldfirm(lambda), give = 1)$`Log-likelihood`}
lam <- seq(from = 0.9, to = 3, len = howmany)
logLike <- sapply(lam, f)
@
%
Thus \fct{f} returns the support at the evaluate, conditional on a
supplied value of $\lambda$.  We may use this to create a profile
likelihood (Figure~\ref{profilesupportlambda}), or to find the maximum
likelihood estimate for $\lambda$:

@@

<<testlambdazero, cache = TRUE, eval = TRUE, echo = FALSE, fig = TRUE>>=
jj <- optimize(f = function(D){f(c(lambda=1, D))}, interval = c(0.1, 0.4), maximum = TRUE)
@
%
Further, we may calculate a support function over the
$(\lambda,D)$-plane, shown in Figure~\ref{contourlambdaD}.  For
example, we may test the hypothesis that $\lambda=1$ [corresponding to
  the left hand side of Figure~\ref{contourlambdaD}] for which the
support is about \Sexpr{round(f(maxv)-jj$objective,2)}, far
exceeding the two units of support per degree of freedom
criteron~\citep{edwards1992}.  Alternatively, we can observe that this
lies in the tail region of its asymptotic null $\chi^2_2$ distribution
with a $p$~value of about \Sexpr{signif(pchisq(f(maxv)-jj$objective,lower.tail=FALSE,df=2),digits=2)}.

@@

As an example, consider Figure~\ref{tikzabcde}.  This shows a partial
probability tree diagram for some of the $5!=120$ possible order
statistics for five competitors.  The standard Plackett-Luce
likelihoods have been modified to account for two groups of mutually
supporting runners: $\left\lbrace a,b,c\right\rbrace$ with support
term $\lambda$, and $\left\lbrace d,e\right\rbrace$ with support term
$\mu$.  From START, the first runner to cross the finishing line has
standard Plackett-Luce probability.  Taking the top path as an
example, we see that the likelihood function for $a\succ b\succ c\succ
d\succ e$ would be
